---
title: "Ames Housing Prices: Part I - EDA & Imputing Missing Data"
author: "Sanjeev Gadre"
date: "June 28, 2019"
output: 
    md_document:
        toc: TRUE
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = TRUE)
```

Loading the required libraries.

```{r libraries, echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)

```

Loading required functions.

```{r functions}
na.count = function (dat){
  dat %>% apply(., 2, is.na) %>% apply(.,2,sum) %>% .[.!=0]
}

```

### Getting the Data

1.  We start by getting a sense of the training data and its structure.
2.  We add a column "SalePrice" to the test data and assign a default value of "1000".
3.  Finally, we combide train and test data, for easier changes to the dataframe structure, imputing missing values and analysis.

```{r get-data}
train = read.csv("./data/train.csv")

head(train)
str(train)

test = read.csv("./data/test.csv")
test$SalePrice = 1000

dat = rbind(train, test)

indx = nrow(train)

```

### Pre-Process - I

1.  We change the data type for a few features to better align it with the type of feature data. Specifically, we convert  `MSSubClass`, `OverallQual`, `OverallCond` from `<int>` to `<fctr>`.
2.  We identify the features in the train subset that report `NA` values

```{r pre-proc-1}
dat$MSSubClass = factor(dat$MSSubClass)
dat$OverallCond = factor(dat$OverallCond)
dat$OverallQual = factor(dat$OverallQual)

na.count(dat[1:indx,])

```

Some features are *incorrectly* coded in that `NA` doesn't mean missing data but has a an alternate meaning. We
convert `NA` to a more appropriate values for the following features:

1.  When a house has *No Alley Access*, it is recorded under the column `Alley` as `NA` which is corrected to   `NoAccess`.
2.  When a house has *No Basement*, it is recorded under the columns `BsmtQual`, `BsmtCond`, `BsmtExposure`, `BsmtFinType1`, `BsmtFinType2` as `NA` which is corrected to `NoBsmt`.
3.  When a house has *No Fireplace*, it is recorded under the column `FireplaceQu` as `NA` which is corrected to `NoFpl`.
4.  When a house has *No Garage*, it is recorded under the columns `GarageType`, `GarageFinish`, `GarageQual`, `GarageCond` as `NA` which is corrected to `NoGrg`.
5.  When a house has *No Pool*, it is recorded under the column `PoolQc` as `NA` which is corrected to        `NoPool`.
6.  When a house has *No Fence*, it is recorded under the column `Fence` as `NA` which is corrected to            `NoFence`.
7.  When a house has *No Miscellaneous Features*, it is recorded under the column `MiscFeature` as `NA` which is corrected to `NoMisc`.

We apply these changes to both train and test subsets.

8.  We identify the features in the train subset that are reporting `NA`.

```{r pre-proc-2}
levels(dat$Alley) = c(levels(dat$Alley), "NoAccess")
dat[is.na(dat$Alley), "Alley"] = "NoAccess"
    
levels(dat$BsmtQual) = c(levels(dat$BsmtQual), "NoBsmt")
dat[is.na(dat$BsmtQual), "BsmtQual"] = "NoBsmt"  
    
levels(dat$BsmtCond) = c(levels(dat$BsmtCond), "NoBsmt")
dat[is.na(dat$BsmtCond), "BsmtCond"] = "NoBsmt"
    
levels(dat$BsmtExposure) = c(levels(dat$BsmtExposure), "NoBsmt")
dat[is.na(dat$BsmtExposure), "BsmtExposure"] = "NoBsmt"
    
levels(dat$BsmtFinType1) = c(levels(dat$BsmtFinType1), "NoBsmt")
dat[is.na(dat$BsmtFinType1), "BsmtFinType1"] = "NoBsmt"
    
levels(dat$BsmtFinType2) = c(levels(dat$BsmtFinType2), "NoBsmt")
dat[is.na(dat$BsmtFinType2), "BsmtFinType2"] = "NoBsmt"
    
levels(dat$FireplaceQu) = c(levels(dat$FireplaceQu), "NoFpl")
dat[is.na(dat$FireplaceQu), "FireplaceQu"] = "NoFpl"
    
levels(dat$GarageType) = c(levels(dat$GarageType), "NoGrg")
dat[is.na(dat$GarageType), "GarageType"] = "NoGrg"
    
levels(dat$GarageFinish) = c(levels(dat$GarageFinish), "NoGrg")
dat[is.na(dat$GarageFinish), "GarageFinish"] = "NoGrg"
    
levels(dat$GarageQual) = c(levels(dat$GarageQual), "NoGrg")
dat[is.na(dat$GarageQual), "GarageQual"] = "NoGrg"
    
levels(dat$GarageCond) = c(levels(dat$GarageCond), "NoGrg")
dat[is.na(dat$GarageCond), "GarageCond"] = "NoGrg"
    
levels(dat$PoolQC) = c(levels(dat$PoolQC), "NoPool")
dat[is.na(dat$PoolQC), "PoolQC"] = "NoPool"
    
levels(dat$Fence) = c(levels(dat$Fence), "NoFence")
dat[is.na(dat$Fence), "Fence"] = "NoFence"
    
levels(dat$MiscFeature) = c(levels(dat$MiscFeature), "NoMisc")
dat[is.na(dat$MiscFeature), "MiscFeature"] = "NoMisc"

na.count(dat[1:indx,])

```

### Exploratory Data Analysis - I

We get additional visibility to the features in the *train* subset reporting `NA` values

```{r eda-1, echo=FALSE}
writeLines("Property Ids reporting NA for GarageYrBlt")
dat[1:indx,] %>% .[is.na(.$GarageYrBlt),c("Id", "GarageType", "GarageYrBlt")]

writeLines("\nDistribution of types of 'Elec. System' used")
table(dat$Electrical[1:indx], dnn = "Elec. System Used") %>% prop.table() %>% round(digits = 4)

writeLines("\nDistribution of types of 'Masonry Veneer' used")
table(dat$MasVnrType[1:indx], dnn = "Masonry Veneer Type") %>% prop.table() %>% round(digits = 4)

writeLines("\nDistribution of 'Masonry Veneer Area'")
hist(dat$MasVnrArea[1:indx], xlab = "Area", main = "Histogram of MasVnrArea")

writeLines("\nDistribution of 'Lot Area' used\n")
hist(dat$LotFrontage[1:indx], xlab = "Area", main = "Histogram of LotArea")

```

*Observations:*

1.  All 81 properties reporting `NA` for `GarageYrBlt` also report that they have no garage. 
2.  The most common value of `Electrical` is `SBrkr` and for `MasVnrType` is `None`
3.  The distribution for `LotFrontage` and `MasVnrArea` is left skewed.

### Pre-Process - II

We use the insights above to impute missing values to both the train and test subset for the 5 features analysed above

1.  We make an experience-based assumption that a property's value is dependent on the presence or absence of a garage rather that the age of the garage. As such, we *drop* the feature `GarageYrBlt` as the information of the presence or absence of a garage is already captured in `GarageType`.
2.  To the `NA` in `Electrical` and `MasVnrType` we impute the mode value of the respective feature based on the remaining train examples.
3.  To the `NA` in `MasVnrArea` and `LotFrontage` we impute the median value of the respective feature based on the remaining train examples.
4.  We verify that there indeed are no more missing values in the **train** subset.
5.  Finally we identify the missing values in the **test** subset

It is important to note that we *only* use the train examples to decide on the median and mode values to impute.

```{r impute-data-1}
dat[is.na(dat$Electrical), "Electrical"] = dat$Electrical[1:indx] %>% table() %>% which.max() %>% names()
dat[is.na(dat$MasVnrType), "MasVnrType"] = dat$MasVnrType[1:indx] %>% table() %>% which.max() %>% names()
dat[is.na(dat$MasVnrArea), "MasVnrArea"] = dat$MasVnrArea[1:indx] %>% median(., na.rm = TRUE)
dat[is.na(dat$LotFrontage), "LotFrontage"] = dat$LotFrontage[1:indx] %>% median(., na.rm = TRUE)

dat = subset(dat, select = -GarageYrBlt)

na.count(dat[1:indx,])

na.count(dat[(indx+1):nrow(dat),])

```

### Exploratory Data Analysis - II

1.  We get some additional visibility to the examples in the test subset reporting `NA`; specifically we ascertain if there are examples reporting `NA` under multiple features.

```{r eda-2}
dat[is.na(dat$MSZoning), c("Id", "Neighborhood", "MSZoning")]
dat[is.na(dat$Utilities), c("Id", "Neighborhood", "Utilities")]
dat[is.na(dat$Exterior1st), c("Id", "MSSubClass", "Exterior1st", "Exterior2nd")]
dat[is.na(dat$BsmtFinSF1), c("Id", "MSSubClass", "BsmtQual", "BsmtFinSF1", "BsmtFinSF2", "BsmtUnfSF", "TotalBsmtSF")]
dat[is.na(dat$BsmtFullBath), c("Id", "MSSubClass", "BsmtQual", "BsmtFullBath", "BsmtHalfBath")]
dat[is.na(dat$KitchenQual), c("Id", "MSSubClass", "KitchenQual")]
dat[is.na(dat$Functional), c("Id", "MSSubClass", "Functional")]
dat[is.na(dat$GarageCars), c("Id", "GarageType", "GarageCars", "GarageArea")]
dat[is.na(dat$SaleType), c("Id", "MSSubClass", "SaleType")]

```

*Observations*

1.  A single exampe 2152 reports `NA` for both `Exterior1st` and `Exterior2nd`. 
2.  Example ids. 2121 and 2189 both report absence of a basement and hence we can impute `0` to all related features like `BsmtFinSF1`, `BsmtFinSF2`, `BsmtUnfSF`, `TotalBsmtSF`, `BsmtFullBath` and `BsmtHalfBath`.
3.  A single example 2577 reports `NA` for both `GarageCars` and `GarageArea`.

*Insights*

1.  We make experience based hypotheses that features:
    a.  `MSZoning` and `Utilities` depend on the `Neighborhood`,
    b.  `Exterior1st`, `Exterior2nd`, `KitchenQual`, `Functional` and `SaleType` depend on the type of house (`MSSubClass`), and 
    c.  `GarageCars` and `GarageArea` depend on typle of Garage (`GarageType`).
2.  To validate this assumption, we look at the distribution of:
    a.  `MSZoning` and `Utilities` for different `Neighborhood`
    b.  `Exterior1st`, `Exterior2nd`, `KitchenQual`, `Functional` and `SaleType` for different `MSSubclass`, and
    c.  `GarageCars` and `GarageArea` for different `GarageType`. 

We look at these distributions only for the train subset examples.

```{r eda-3, echo=FALSE}
writeLines("The first table of each pair of tables below show the overall distribution of a feature across all examples and the second one the distribution of the feature stratified by 'MSSubClass'\n")

table(MSZoning = dat$MSZoning[1:indx], dnn = "MSZoning") %>% prop.table() %>% round(digits = 4)
table(MSZoning = dat$MSZoning[1:indx], Neighborhood = dat$Neighborhood[1:indx]) %>% 
    prop.table(margin = 2) %>% round(digits = 4)
writeLines("\n")
table(Utilities = dat$Utilities[1:indx], dnn = "Utilities") %>% prop.table() %>% round(digits = 4)
table(Utilities = dat$Utilities[1:indx], Neighborhood = dat$Neighborhood[1:indx]) %>% 
    prop.table(margin = 2) %>% round(digits = 4)
writeLines("\n")
table(Xterior1 = dat$Exterior1st[1:indx], dnn = "Exterior1st") %>% prop.table() %>% round(digits = 4)
table(Xterior1 = dat$Exterior1st[1:indx], MSSubclass = dat$MSSubClass[1:indx]) %>% 
    prop.table(margin = 2) %>% round(digits = 4)
writeLines("\n")
table(Xterior2 = dat$Exterior2nd[1:indx], dnn = "Exterior2nd") %>% prop.table() %>% round(digits = 4)
table(Xterior2 = dat$Exterior2nd[1:indx], MSSubclass = dat$MSSubClass[1:indx]) %>% 
    prop.table(margin = 2) %>% round(digits = 4)
writeLines("\n")
table(KitchenQual = dat$KitchenQual[1:indx], dnn = "KitchenQual") %>% prop.table() %>% round(digits = 4)
table(KitchenQual = dat$KitchenQual[1:indx], MSSubclass = dat$MSSubClass[1:indx]) %>% 
    prop.table(margin = 2) %>% round(digits = 4)
writeLines("\n")
table(Functional = dat$Functional[1:indx], dnn = "Functional") %>% prop.table() %>% round(digits = 4)
table(Functional = dat$Functional[1:indx], MSSubclass = dat$MSSubClass[1:indx]) %>% 
    prop.table(margin = 2) %>% round(digits = 4)
writeLines("\n")
table(SaleType = dat$SaleType[1:indx], dnn = "SaleType") %>% prop.table() %>% round(digits = 4)
table(SaleType = dat$SaleType[1:indx], MSSubclass = dat$MSSubClass[1:indx]) %>% 
    prop.table(margin = 2) %>% round(digits = 4)
writeLines("\n")
table(GarageCars = dat$GarageCars[1:indx], dnn = "GarageCars") %>% prop.table() %>% round(digits = 4)
table(GarageCars = dat$GarageCars[1:indx], GarageType = dat$GarageType[1:indx]) %>% 
    prop.table(margin = 2) %>% round(digits = 4)

garage.plot.1 = dat[1:indx,] %>% filter(.$GarageType != "NoGrg") %>% 
    ggplot(aes(x = GarageArea))+ geom_density()+ 
    labs(title = "Overall Garage Area Distribution", x = "Area in sq.ft", y = "")
garage.plot.2 = dat[1:indx,] %>% filter(.$GarageType != "NoGrg") %>% ggplot(aes(x = GarageArea))+ 
    geom_density()+ facet_grid(GarageType~.)+ 
    labs(title = "Area Distribution by Garage Type", x = "Area in sq.ft", y = "")
grid.arrange(garage.plot.1, garage.plot.2, ncol = 2)

```

1.  The tables and the graphs above validate our hypotheses that, features - 
    a.  `MSZoning` and `Utilities` depend on `Neighborhood`
    b.  `Exterior1st`, `Exterior2nd`, `KitchenQual`, `Functional` and `SaleType` depend on the type of house (`MSSubClass`), and 
    c.  `GarageCars` and `GarageArea` depend on typle of Garage (`GarageType`).

### Impute Missing Data

1.  We impute 
    a.  the mode values for missing `MSZoning` and `Utilities` of appropriate `Neighborhood`
    b.  the mode values for missing `Exterior1st`, `Exterior2nd`, `KitchenQual`, `Functional` and `SaleType` of appropriate `MSSubClass`, and 
    c.  the mode and median values respectively for missing `GarageCars` and `GarageArea` of appropriate `GarageType`
2.  Example ids. 2121 and 2189 both report absence of a basement and hence we can impute `0` to all related features like `BsmtFinSF1`, `BsmtFinSF2`, `BsmtUnfSF`, `TotalBsmtSF`, `BsmtFullBath` and `BsmtHalfBath`.
3.  Finally, we reconfirm that there are no more missing values.

```{r impute-data-2}
#   MSZoning
x = dat$MSZoning[dat$Neighborhood == "IDOTRR"] %>% table() %>% which.max() %>% names()
dat$MSZoning[c(1916, 2217, 2251)] = x
x = dat$MSZoning[dat$Neighborhood == "Mitchel"] %>% table() %>% which.max() %>% names()
dat$MSZoning[2905] = x
#   Utilities
x = dat$Utilities[dat$Neighborhood == "IDOTRR"] %>% table() %>% which.max() %>% names()
dat$Utilities[1916] = x
x = dat$Utilities[dat$Neighborhood == "Gilbert"] %>% table() %>% which.max() %>% names()
dat$Utilities[1946] = x
#   Exterior1st
x = dat$Exterior1st[dat$MSSubClass == 30] %>% table() %>% which.max() %>% names()
dat$Exterior1st[2152] = x
#   Exterior2nd
x = dat$Exterior2nd[dat$MSSubClass == 30] %>% table() %>% which.max() %>% names()
dat$Exterior2nd[2152] = x
#   KitchenQual
x = dat$KitchenQual[dat$MSSubClass == 50] %>% table() %>% which.max() %>% names()
dat$KitchenQual[1556] = x
#   Functional
x = dat$Functional[dat$MSSubClass == 20] %>% table() %>% which.max() %>% names()
dat$Functional[2217] = x
x = dat$Functional[dat$MSSubClass == 50] %>% table() %>% which.max() %>% names()
dat$Functional[2474] = x
#   SaleType
x = dat$SaleType[dat$MSSubClass == 20] %>% table() %>% which.max() %>% names()
dat$SaleType[2490] = x
#   GarageCars
x = dat$GarageCars[dat$GarageType == "Detchd"] %>% median(na.rm = TRUE)
dat$GarageCars[2577] = x
#   GarageArea
x = dat$GarageArea[dat$GarageType == "Detchd"] %>% median(na.rm = TRUE)
dat$GarageArea[2577] = x
#   BsmtFinSF1, BsmtFinSF2, BsmtUnfSF, TotalBsmtSF, BsmtFullBath and BsmtHalfBath.
dat[c(2121, 2189), c("BsmtFinSF1", "BsmtFinSF2", "BsmtUnfSF", "TotalBsmtSF", "BsmtFullBath", "BsmtHalfBath")] =
    0

na.count(dat[(indx+1):nrow(dat),])

```

### Saving a Baseline

1.  We have "cleaned" both the train and test subsets.
2.  We assign the `Id` as rownames for the examples and then drop the `Id` feature.
3.  Finally, we separate the train and the test subsets and save then as *.Rda* files as baseline versions

```{r baseline}
rownames(dat) = dat$Id
dat = subset(dat, select = -Id)
train = dat[1:indx,]
test = dat[(indx+1):nrow(dat),]

saveRDS(train, "./RDA/train")
saveRDS(test, "./RDA/test")

```

